---
title: "MindTheGap"
author: "Callum Hollywood"
date: "28 December 2017"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

packs <- c("tidyverse", "dplyr", "gapminder", "reshape2", "rworldmap")
lapply(packs, require, character.only = T)
options(scipen = 999)
```

## A Brief Exploratory Data Analysis of the Gapminder Dataset
## Shipped with the R Gapminder Package
***
***
***
<br>

#### We can get a feel for the general composition of this dataset
#### by looking at both ends of the dataframe, as it is shipped. Supported by some
#### additional basic descriptive statistics: ...

<br>

```{r cars}
head(gapminder)
tail(gapminder)
str(gapminder)
summary(gapminder)
```
<br>

#### Data Summary: -- 
#### It looks like we have over 1700 observations detailing measurements
#### for 3 key features, 1) Life Expectancy, 2) Population and 3) GDP per Capita
#### for 142 countries, which have also been labeled with their appropriate continent.
#### The duration for capturing this data appears to be from 1952 to 2007 and has been captured 
#### at 5 year intervals.

<br>

***
***
***

<br>

#### We can proceed to examine the gapminder features, by country, over time: ...

<br>
```{r features1, fig.align="center", echo=TRUE}
#========================
# plot lifeExp over the survey duration, by country 
#========================

ggplot(gapminder, aes(x = year, y = lifeExp, group = country)) +
  geom_line() +
  theme(legend.position = "none") +
  ggtitle("Life Expectancy by Country") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(data = gapminder[gapminder$year == 2007,], aes(label = country)) +
  xlab("Year") +
  ylab("Life Expectancy (Years)") 
```
<br>

#### Looks like there's a lot going on with all the countries represented above.
#### What we can discern though is that there appears to be a general increase in
#### life expectancy, though there are definitely some countries that seem to buck that trend.
#### We will return to this and examine in greater depth!
***
***
***

<br>

```{r features2, fig.align="center", echo=TRUE}

#========================
# plot pop over the survey duration, by country 
#========================

ggplot(gapminder, aes(year, pop, group = country)) +
  geom_line() +
  theme(legend.position = "none") +
  ggtitle("Population by Country") +
  geom_text(data = gapminder[gapminder$year == 2007,], aes(label = country)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Year") +
  ylab("Population")

```

<br>

#### Jeez Louise - Look at the population climbs by China and India!
***
***
***

<br>

```{r features3, fig.align="center", echo=TRUE}

#========================
# plot gdpPercap over the survey duration, by country 
#========================

ggplot(gapminder, aes(year, gdpPercap, group = country)) +
  geom_line() +
  theme(legend.position = "none") +
  ggtitle("All Countries GDP Per Capita") +
  geom_text(data = gapminder[gapminder$year == 1962,], aes(label = country)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Year") +
  ylab("GDP Per Capita ($)")
```

<br>

#### In generel, i.e. omitting the GDP per Capita of Kuwait, we can see that this feature
#### tends to be increasing over time.  It would be interesting to see what is going on with Kuwait later. 
#### And if there are any other countries that have a reduced GDP
***
***
***
#### The graphs above show the gapminder features at the country level, and it all got
#### a bit busy on the eyes. Let's aggregate to the 'Continent' level: ...

<br>

```{r continents, fig.align="center", echo=TRUE}

continent_view <- gapminder %>% 
  group_by(continent, year) %>% 
  summarise(mean_lifeExp = mean(lifeExp, na.rm = T),
            mean_pop = mean(pop, na.rm = T),
            mean_gdpPercap = mean(gdpPercap, na.rm = T))

head(continent_view)

```

```{r continent1, fig.align="center", echo=TRUE}

#========================
# plot mean_lifeExp over the survey duration, by continent 
#========================

continent_view %>% 
  select(-c(mean_pop, mean_gdpPercap)) %>% 
  ggplot(aes(year, mean_lifeExp, group = continent)) +
  geom_line() +
  ggtitle("Average Life Expectancy by Continent") +
  geom_text(data = continent_view %>% filter(year == 2002) ,
            aes(label = continent))+
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Years") +
  ylab("Average Life Expectancy (Years)")

```

<br>

#### Across the board we can see that life expecnacy is on the increase and that there appears to 
#### be a distinct difference in average life expectancy across the continents. Of particular note 
#### is that not ony is Africa trailing, it appears to have levelled off through the late 80's and 90's.
***
***
***

<br>

```{r continent2, fig.align="center", echo=TRUE}

#========================
# plot mean_pop over the survey duration, by continent 
#========================

continent_view %>% 
  select(-c(mean_lifeExp, mean_gdpPercap)) %>% 
  ggplot(aes(year, mean_pop, group = continent)) +
  geom_line() +
  ggtitle("Average Population by Continent") +
  geom_text(data = continent_view %>% filter(year == 2002),
            aes(label = continent)) +
  xlab("Years") +
  ylab("Population") +
  theme(plot.title = element_text(hjust = 0.5))

```

<br>

***
***
***

<br>

```{r continent3, fig.align="center", echo=TRUE}

#========================
# plot mean_gdpPercap over the survey duration, by continent 
#========================

continent_view %>% 
  select(-c(mean_lifeExp, mean_pop)) %>% 
  ggplot(aes(year, mean_gdpPercap, group = continent)) +
  geom_line() +
  ggtitle("Average GDP per Capita by Continent") +
  geom_text(data = continent_view %>% filter(year == 2002),
            aes(label = continent)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Year") +
  ylab("Average GDP ($)")
```

<br>

#### It appears that the average GDP in both Oceania and Eurpoe is outpacing
#### Asia and the Americas, which are all pulling away from Africa
#### Africa definitely seems to out of step with the other continents.
***
***
***
#### Let's have a look at the continents individually to see how the gapminder features are interplaying:...  

<br>

```{r africa, fig.align="center", out.width="2000px", out.height="1000px", echo=TRUE}

africa <- gapminder %>% filter(continent == "Africa") %>% select(-c(continent))

africa_melt <- melt(africa, 
                    id.vars = c("country", "year"),
                    variable.name = "feature",
                    value.name = "value") 

africa_melt %>% 
  ggplot(aes(year, value, group = country)) +
  geom_line() +
  facet_wrap(~ feature, scales = "free_y") +
  geom_text(data = africa_melt %>% filter(year == 1972),
            aes(label = country)) +
  ggtitle("Africa; Life Expectancy, Population and GDP per Capita") +
  theme(plot.title = element_text(hjust = 0.5))

```

<br>

***
***
***

<br>

```{r americas, fig.align="center", out.width="2000px", out.height="1000px", echo=TRUE}

americas <- gapminder %>% filter(continent == "Americas") %>%  select(-c(continent))

americas_melt <- melt(americas, 
                      id.vars = c("country", "year"),
                      variable.name = 'feature',
                      value.name = "value")

americas_melt %>% 
  ggplot(aes(year, value, group = country)) +
  geom_line() +
  facet_wrap(~feature, scales = 'free_y') +
  geom_text(data = americas_melt %>% filter(year == 1997),
            aes(label = country)) +
  ggtitle("Americas; Life Expectancy, Population and GDP per Capita") +
  theme(plot.title = element_text(hjust = 0.5))

```

<br>

***
***
***

<br>


```{r asia, fig.align="center", out.width="2000px", out.height="1000px", echo=TRUE}

asia <- gapminder %>% filter(continent == "Asia") %>% select(-c(continent))

asia_melt <- melt(asia, 
                  id.vars = c("country", "year"),
                  variable.name = "feature",
                  value.name = "value")

asia_melt %>% 
  ggplot(aes(year, value, group = country)) +
  geom_line() +
  facet_wrap(~ feature, scales = "free_y") +
  geom_text(data = asia_melt %>% filter(year == 1997),
            aes(label = country)) +
  ggtitle("Asia; Life Expectancy, Population and GDP per Capita") +
  theme(plot.title = element_text(hjust = 0.5))

```

<br>

***
***
***

<br>

```{r europe, fig.align="center", out.width="2000px", out.height="1000px", echo=TRUE}
europe <- gapminder %>% filter(continent == "Europe") %>% select(-c(continent))

europe_melt <- melt(europe,
                    id.vars = c("country", "year"),
                    variable.name = "feature",
                    value.name = "value")


europe_melt %>% 
  ggplot(aes(year, value, group = country)) +
  geom_line() +
  facet_wrap(~ feature, scales = 'free_y') +
  geom_text(data = europe_melt %>% filter(year == 1997),
            aes(label = country)) +
  ggtitle("Europe; Life Expectancy, Population and GDP per Capita") +
  theme(plot.title = element_text(hjust = 0.5))

```

<br>

***
***
***

<br>

```{r oceania, fig.align="center", out.width="2000px", out.height="1000px", echo=TRUE}

oceania <- gapminder %>% filter(continent == "Oceania") %>% select(-c(continent))

oceania_melt <- melt(oceania,
                    id.vars = c("country", "year"),
                    variable.name = "feature",
                    value.name = "value")

oceania_melt %>% 
  ggplot(aes(year, value, group = country)) +
  geom_line() +
  facet_wrap(~ feature, scales = "free_y") +
  geom_text(data = oceania_melt %>% filter(year == 1997),
          aes(label = country)) +
  ggtitle("Oceania; Life Expectancy, Population and GDP per Capita") +
  theme(plot.title = element_text(hjust = 0.5))

```

<br>

***
***
***

<br>

#### Africa still requires further examination. Let's have a look at all the countries where there was a dip in life expectancy observed, at some/any point in the data:...

<br>

```{r africa_le_dip, fig.align="center", out.width="2000px", out.height="1000px", echo=TRUE}

##################################################################
# Drill into Africa and examine at the country level
# Filter out countries that never took a dip in lifeExp at any point
# Facet for better visuals
##################################################################

lifeexpdrops <- gapminder::gapminder %>% 
  filter(continent == "Africa") %>% 
  select(-c(continent, pop, gdpPercap)) %>% 
  spread(year, lifeExp)

indx <- colSums(apply(lifeexpdrops %>% select(-c(country)), 1L, diff) > 0L) == (ncol(lifeexpdrops%>% select(-c(country))) - 1L)
lifeexpdrops <-  lifeexpdrops[!indx,]$country

gapminder::gapminder %>% 
  filter(continent == "Africa") %>% 
  filter(country %in% lifeexpdrops) %>% 
  select(-c(continent)) %>% 
  ggplot(aes(year, lifeExp, group = country)) +
  geom_line() +
  facet_wrap(~ country) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("African Countries that Exhibited at Least One Life Expectatncy Reduction") +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Year") +
  ylab("Life Expectancy (Years)")

```

<br>

***
***
***

<br>

#### Let's start to visualise with a map or two:...

<br>

```{r africa_le_drop, out.width="2000px", out.height="1000px", echo=TRUE}


lifeexpdrops <- gapminder::gapminder %>% 
  filter(continent == "Africa") %>% 
  select(-c(continent, pop, gdpPercap)) %>% 
  spread(year, lifeExp)

indx <- colSums(apply(lifeexpdrops %>% select(-c(country)), 1L, diff) > 0L) == (ncol(lifeexpdrops%>% select(-c(country))) - 1L)
lifeexpdrops <-  lifeexpdrops[!indx,]$country

africa <- gapminder::gapminder %>% 
  filter(continent == "Africa") %>% 
  select(-c(continent)) %>% 
  filter(year == 2002) %>% 
  mutate(lifeExpDrop = if_else(country %in% lifeexpdrops, "dropped at least once", "has never dropped"))

sPDF <- joinCountryData2Map(africa,
                            joinCode = "NAME",
                            nameJoinColumn = "country")

par(mai = c(0,0,0.2,0), xaxs="i", yaxs = "i")
mapCountryData(sPDF, 
               nameColumnToPlot = "lifeExpDrop", 
               catMethod="categorical", 
               numCats = 2,
               addLegend =T,
               mapTitle = "African Countries with at least One Life Expectancy Drop",
               mapRegion = "Africa")

country_coord<-data.frame(coordinates(sPDF),stringsAsFactors=F)
country_coord_B <- country_coord[row.names(country_coord) %in% africa$country,]

text(x = country_coord_B$X1, y = country_coord_B$X2, 
     labels = row.names(country_coord_B))

```

<br>

***
***
***

<br>

```{r the_rest_le_drop, out.width="2000px", out.height="1000px", echo=TRUE}


lifeexpdrops <- gapminder::gapminder %>% 
  select(-c(continent, pop, gdpPercap)) %>% 
  spread(year, lifeExp)

indx <- colSums(apply(lifeexpdrops %>% select(-c(country)), 1L, diff) > 0L) == (ncol(lifeexpdrops%>% select(-c(country))) - 1L)
lifeexpdrops <-  lifeexpdrops[!indx,]$country

worldLifeExpDrop <- gapminder::gapminder %>% 
  select(-c(continent)) %>% 
  filter(year == 2002) %>% 
  mutate(lifeExpDrop = if_else(country %in% lifeexpdrops, "dropped at least once", "has never dropped"))


worldLifeExpDropExcptAfrica <- worldLifeExpDrop %>% 
  filter(lifeExpDrop == "dropped at least once") %>% 
  filter(!country %in% africa$country)

sPDF_World <- joinCountryData2Map(worldLifeExpDrop,
                            joinCode = "NAME",
                            nameJoinColumn = "country")

par(mai = c(0,0,0.2,0), xaxs="i", yaxs = "i")
mapCountryData(sPDF_World, 
               nameColumnToPlot = "lifeExpDrop", 
               catMethod="categorical", 
               numCats = 2,
               addLegend =T,
               mapTitle = "Countries with at least One Life Expectancy Drop"
               )

#country_coord <- data.frame(coordinates(sPDF_World),stringsAsFactors=F)
country_coord_C <- country_coord[row.names(country_coord) %in% worldLifeExpDropExcptAfrica$country,]

text(x = country_coord_C$X1, y = country_coord_C$X2, 
     labels = row.names(country_coord_C))

```

<br>

***
***
***

<br>

```{r pop_drop, out.width="2000px", out.height="1000px", echo=TRUE}

pop_drops <- gapminder::gapminder %>% 
  select(-c(continent, lifeExp, gdpPercap)) %>% 
  spread(year, pop)

indx <- colSums(apply(pop_drops %>% select(-c(country)), 1L, diff) > 0L) == (ncol(pop_drops%>% select(-c(country))) - 1L)
pop_drops <-  pop_drops[!indx,]$country

world_popdrops <- gapminder::gapminder %>% 
  select(-c(continent)) %>% 
  filter(year == 2002) %>% 
  mutate(pop_drops = if_else(country %in% pop_drops, "dropped at least once", "has never dropped"))

world_popdrops_slim <- world_popdrops[world_popdrops$pop_drops == 'dropped at least once',]

sPDF_worldpopdrop <- joinCountryData2Map(world_popdrops,
                                  joinCode = "NAME",
                                  nameJoinColumn = "country")

par(mai = c(0,0,0.2,0), xaxs="i", yaxs = "i")
mapCountryData(sPDF_worldpopdrop, 
               nameColumnToPlot = "pop_drops", 
               catMethod="categorical", 
               numCats = 2,
               addLegend =T,
               mapTitle = "Countries with at least One Population Drop")

#country_coord <- data.frame(coordinates(sPDF_World),stringsAsFactors=F)
country_coord_D <- country_coord[row.names(country_coord) %in% world_popdrops_slim$country,]

text(x = country_coord_D$X1, y = country_coord_D$X2, 
     labels = row.names(country_coord_D))

```

<br>

***
***
***

<br>

```{r gdp_drop, out.width="2000px", out.height="1000px", echo=TRUE}


gdp_drops <- gapminder::gapminder %>% 
  select(-c(continent, lifeExp, pop)) %>% 
  spread(year, gdpPercap)

indx <- colSums(apply(gdp_drops %>% select(-c(country)), 1L, diff) > 0L) == (ncol(gdp_drops%>% select(-c(country))) - 1L)
gdp_drops <-  gdp_drops[!indx,]$country

world_gdpdrops <- gapminder::gapminder %>% 
  select(-c(continent)) %>% 
  filter(year == 2002) %>% 
  mutate(gdp_drops = if_else(country %in% gdp_drops, "dropped at least once", "has never dropped"))

world_gdpdrops_slim <- world_gdpdrops[world_gdpdrops$gdp_drops == 'dropped at least once',]


sPDF_worldgdpdrop <- joinCountryData2Map(world_gdpdrops,
                                         joinCode = "NAME",
                                         nameJoinColumn = "country")

par(mai = c(0,0,0.2,0), xaxs="i", yaxs = "i")
mapCountryData(sPDF_worldgdpdrop, 
               nameColumnToPlot = "gdp_drops", 
               catMethod="categorical", 
               numCats = 2,
               addLegend =T,
               mapTitle = "Countries with at least One GDP Drop")

#country_coord <- data.frame(coordinates(sPDF_World),stringsAsFactors=F)
country_coord_E <- country_coord[!row.names(country_coord) %in% world_gdpdrops_slim$country,]

text(x = country_coord_E$X1, y = country_coord_E$X2, 
     labels = row.names(country_coord_E))

```

<br>

#### It appears that with respect to GDP it is not unusual to have a dip in GDP year on year,
#### for many countries. Let's instead have a look to see if any countries have GDPs in 2007 that 
#### are less than our starting point in 1952: ...

<br>

```{r total_gdp_drop, fig.align="center", out.width="2000px", out.height="1000px", echo=TRUE}

gdp_world_drops <- gapminder::gapminder %>% 
  filter(year %in% c(1952, 2007)) %>% 
  select(-c(lifeExp, pop, continent)) %>% 
  spread(year, gdpPercap) %>% 
  mutate(gdp_reduced = `2007` - `1952`) %>% 
  mutate(gdp_reduced_cat = if_else(gdp_reduced < 0, "Has Reduced Overall", "Has Increased Overall")) %>% 
  filter(gdp_reduced_cat == "Has Reduced Overall")

gdp_world_drops %>% select(-c(gdp_reduced_cat))

sPDF_gdp_world_drops <- joinCountryData2Map(gdp_world_drops,
                                            joinCode = "NAME",
                                            nameJoinColumn = "country")

par(mai = c(0,0,0.2,0), xaxs="i", yaxs = "i")
mapCountryData(sPDF_gdp_world_drops, 
               nameColumnToPlot = "gdp_reduced_cat", 
               catMethod="categorical", 
               numCats = 2,
               addLegend =F,
               mapTitle = "Countries with a Smaller GDP in 2007 than in 1952"
)

#country_coord <- data.frame(coordinates(sPDF_gdp_world_drops),stringsAsFactors=F)
country_coord_F <- country_coord[row.names(country_coord) %in% sPDF_gdp_world_drops$country,]

text(x = country_coord_F$X1, y = country_coord_F$X2, 
     labels = row.names(country_coord_F))

```